Predator clusters

### 2023-02-22 reports/04_pred_clus.qmd
Cluster comparison

library(ggtree)
library(sf)
library(targets)
library(tidyverse)

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_stations %>%
  group_by(amlr.station, species) %>% 
  summarize(count = sum(count), .groups = "drop") %>%
  group_by(species) %>%
  summarize(stations_present = n(),
            station_frac = stations_present / total_stations,
            .groups = "drop") %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)

@@

Hierarchical clustering of stations by predator community.

```{r}
#| label: predator_clusters

clus <- cutree(sightings_clust, 4)
g <- split(names(clus), clus)
p <- ggtree(sightings_clust)
station_clades <- sapply(g, function(n) MRCA(p, n))
p <- groupClade(p, station_clades, group_name = "subtree") + 
  aes(color = subtree)

p +
  layout_dendrogram() +
  theme_dendrogram()

@@

>>>>

library(ggmosaic)
library(ggtree)
library(sf)
library(targets)
library(tidyverse)

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "zoop"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

total_stations <- n_distinct(zoop$amlr.station)
pred_frac <- predators_stations %>%
  group_by(amlr.station, species) %>% 
  summarize(count = sum(count), .groups = "drop") %>%
  group_by(species) %>%
  summarize(stations_present = n(),
            station_frac = stations_present / total_stations,
            .groups = "drop") %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)

@@

Hierarchical clustering of stations by predator community. Chose 5 clusters a priori to match Dietrich et al. (2021). Note: `r nrow(zoop)` stations were included in the zooplankton clustering, but only `r length(sightings_clust$labels)` stations were used for predator clustering.

```{r}
#| label: predator_dendro

# Cut predators into five clusters
clust <- cutree(sightings_clust, 5)
g <- split(names(clust), clust)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

p %>% 
  groupClade(pred_clust_mrca, group_name = "Predator cluster") + 
  aes(color = `Predator cluster`) +
  layout_dendrogram() +
  theme_dendrogram()
```

How do the predator clusters compare to the zooplankton clusters?

```{r}
#| label: predator_mosaic
#| warning: false

# Join with zooplankton clusters
both_clust <- zoop %>% 
  transmute(amlr.station, 
            zoop_clust = factor(`Winter Cluster factor`)) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

# Mosaic plot
cluster_pal = c(`1` = "#96B687",
                `2a` = "#9CB2C8",
                `2b` = "#CAB3CC",
                `3a` = "#D29A97",
                `3b` = "#E8AD5C")
ggplot(both_clust) +
  geom_mosaic(aes(x = product(zoop_clust, pred_clust), 
                  fill = zoop_clust)) +
  geom_mosaic_text(aes(x = product(zoop_clust, pred_clust),
                       label = after_stat(.wt)),
                   as.label = TRUE) +
  scale_fill_manual(values = cluster_pal) +
  labs(x = "Predator cluster",
       y = "Zooplankton cluster",
       fill = "Zooplankton cluster") +
  theme_classic()
```

The above mosaic plot shows the distribution of stations by predator and zooplankton stations. For example, the bottom left rectangle indicates 9 stations were assigned to zooplankton cluster 1 and predator cluster 1. Rectangles color-coded by zooplankton cluster using the color palette from Dietrich et al. 2021 Fig 3a.

Upon visual inspection of the mosaic plot, I don't perceive any clear correlations between the zooplankton and predator cluster assignments. I tested the strength of the relationship between the two cluster assignments through simulation of the Rand index (RI). RI is a measure of similarity between two clusterings ranging from 0 (no agreement) to 1 (total agreement). The histogram below shows the distribution of RI values when the cluster assignments were randomly shuffled 100,000 times. The vertical red line is the observed RI value.

```{r}
#| label: randindex

set.seed(1311)

expected_ri <- replicate(
  100000,
  aricode::RI(
    sample(both_clust$zoop_clust, nrow(both_clust)),
    sample(both_clust$pred_clust, nrow(both_clust))
  )
)

observed_ri <- aricode::RI(both_clust$zoop_clust, both_clust$pred_clust)

ri_pval <- sum(expected_ri > observed_ri) / 100000

ggplot(tibble(`Rand index` = expected_ri), aes(`Rand index`)) +
  geom_histogram(binwidth = 0.001) +
  geom_vline(xintercept = observed_ri, color = "red") +
  theme_classic()

@@

Based on this simulation, the correlation between the zooplankton and predator assignments *is* significant (p = `r as.character(ri_pval)`). But I haven't figured out how to assess effect size yet, which I imagine is small.
